---
title: "IDS investigation worksheet"
author: "by Pipeline: Naomi Pihlaja, Jiaxi Yan, Yupeng Qiu, Cici Chen"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
# Add any other libraries here



```


```{r load-data}
# load your data 
data <- read_csv("AVONET3_BirdTree.csv") %>% 
  view()



```
```{r beak length and depth correlation}
#Hypothesis: Bird beak depth increases as beak length increases. Positive correlation.

ggplot(data, aes(x=Beak.Width, y=Beak.Depth, color = Centroid.Latitude))+
         geom_point()+
  geom_smooth(method= "lm")+
  labs(title = "Bird beak width & depth",
       x = "Beak width (mm)", 
       y = "Beak depth (mm)")



#Hypothesis confirmed to be true.
  
```

```{r latitude and winglength}

#fit a linear regression model to predict latitude based on wing length

```


```{r habitat exploration}
data %>% 
  group_by(Habitat) %>% 
  count() %>% 
  arrange(desc(n)) %>% 
  view()

```
