---
title: "Project title"
author: "by Team-Name: Naomi Pihlaja, Jiaxi Yan, Yupeng Qiu, J|iayi Chen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
library(tidymodels)



```


```{r load-data, include=FALSE}
data <- read_csv("data/AVONET3_BirdTree.csv")

```


## Research Question

Can we accurately predict certain characteristics of birds using chosen explanatory variables? If we can, we could predict lifestyle elements of lesser known, lesser investigated and/ or bird species. These predictions could be used in scientific research for wildlife conservation and investigation.


## Data

Avonet birdtree contains summary statistics on specific characteristics of birds. Many are numerical variables (eg. Beak depth), but some are categorical (eg. Trophic level).Over 10,000 observations allows for a big sample size.

```{r labelling variables}

tibble(
  variable = c("Centroid Latitude", "Trophic level", "Beak width", "Beak depth", "Habitat", "Mass", "Wing length", "Hangwingindex"),
  Type = c("Numerical", "Categorical", "Numerical", "Numerical", "Categorical", "Numerical", "Numerical",  "Numerical"),
  Description = c("The geometric centre of the species range", "Categorical variable based on the composition of a birds' diet", "Width of the beak at the anterior edge of the nostrils", "Depth of the beak at the anterior edge of the nostrils", "Where the bird lives", "Body mass of species average", "Length from the carpal joint to the tip of the longest primary on the unflattened wing", "A measurement of a bird's wing shape that indicates flight efficiency and dispersal ability")
)
```

 
figshare.com. (n.d.). AVONET: morphological, ecological and geographical data for all birds (Tobias et al 2022 Ecology Letters doi: https://doi.org/10.111/ele.13898). [online] Available at: https://figshare.com/s/b990722d72a26b5bfead. 





## Findings

The write-up of your project and findings go here. 

Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation. This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work. Feel free to split this section into subsections to make it easier to read. 

The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.





## References

List any references here. You should, at a minimum, list your data source.


