---
title: "Project title"
author: "by Team-Name: Naomi Pihlaja, Jiaxi Yan, Yupeng Qiu, J|iayi Chen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
library(tidymodels)



```


```{r load-data, include=FALSE}
data <- read_csv("data/AVONET3_BirdTree.csv")

```


## Research Question

Can we accurately predict certain characteristics of birds using chosen explanatory variables? If we can, we could predict lifestyle elements of lesser known, lesser investigated and/ or bird species. These predictions could be used in scientific research for wildlife conservation and investigation.


## Data

figshare.com. (n.d.). AVONET: morphological, ecological and geographical data for all birds (Tobias et al 2022 Ecology Letters doi: https://doi.org/10.111/ele.13898). [online] Available at: https://figshare.com/s/b990722d72a26b5bfead. 





## Findings

#Introduction

Bird morphology and ecological traits offer valuable insight into how different species adapt to their environments. The AVONET dataset contains thousands of bird species with detailed measurements of their morphology, habitat, and trophic ecology. This rich dataset allows us to explore questions regarding trait correlations and predictive modelling.

Our analysis investigates two key questions inspired by our initial exploration:

1. To what extent can morphological traits explain ecological characteristic such as trophic level or latitude?
2. Which morphological variables show meaningful correlations with each other?

These questions were motivated by observations and hypothesis generated during our group presentation

#Data Overview

Avonet birdtree contains summary statistics on specific characteristics of birds. Many are numerical variables (eg. Beak depth), but some are categorical (eg. Trophic level).Over 10,000 observations allows for a big sample size.

```{r labelling variables 1, echo=FALSE}

tibble(
  variable = c("Centroid Latitude", "Trophic level", "Beak width", "Beak depth", "Habitat", "Mass", "Wing length", "Hangwingindex"),
  Type = c("Numerical", "Categorical", "Numerical", "Numerical", "Categorical", "Numerical", "Numerical",  "Numerical"),
  Description = c("The geometric centre of the species range", "Categorical variable based on the composition of a birds' diet", "Width of the beak at the anterior edge of the nostrils", "Depth of the beak at the anterior edge of the nostrils", "Where the bird lives", "Body mass of species average", "Length from the carpal joint to the tip of the longest primary on the unflattened wing", "A measurement of a bird's wing shape that indicates flight efficiency and dispersal ability")
)
```


# Exploratory Statistics
1. Correlation: Beak Width & Beak Depth
We began by exploring correlations between traits we suspecteed to be biologically related. A scatterplot with a linear trend line indicated a strong positive correlation between beak width and beak depth.

```{r beak length and depth correlation, echo=FALSE}
#Hypothesis: Bird beak depth increases as beak length increases. Positive correlation.

ggplot(data, aes(x=Beak.Width, y=Beak.Depth, color = Centroid.Latitude))+
         geom_point()+
  geom_smooth(method= "lm")+
  labs(title = "Bird beak width & depth",
       x = "Beak width (mm)", 
       y = "Beak depth (mm)")

```

2. Predicting Latitude
We attempted to predict centroid latitude using morphological variables. However, our model showed no meaningful predictive signal. This suggests that global latitude does not strongly shape overall morphological configurations once phylogenetically diverse species are pooled.
```{r predicting latitude, echo=FALSE}
# Choose needed data
data_lat1 <- data %>% 
  select(Beak.Length_Culmen, 
         Beak.Length_Nares, 
         Beak.Depth, 
         Range.Size, 
         Habitat, 
         Mass, 
         Centroid.Latitude)
# Modify col names
colnames(data_lat1) <- c("Beak_length_C", 
                        "Beak_Length_N", 
                        "Beak_Depth", 
                        "Range_Size", 
                        "Habitat", 
                        "Mass", 
                        "Latitude")


# Start modelling
set.seed(2801)
data_split1 <- initial_split(data_lat1, 0.8)
data_train1 <- training(data_split1)
data_test1 <- testing(data_split1)


# Construct workflow
data_rec1 <- recipe(Latitude ~ ., data = data_lat1) %>% 
  step_naomit(all_predictors()) %>% 
  step_dummy(all_nominal(), -all_outcomes())

data_mod1 <- linear_reg() %>% 
  set_engine("lm")

data_wflow1 <- workflow() %>% 
  add_model(data_mod1) %>% 
  add_recipe(data_rec1)

data_fit1 <- data_wflow1 %>% 
  fit(data = data_train1)
# tidy(data_fit)

data_pred1 <- predict(data_fit1, data_test1) %>% 
  bind_cols(data_test1) %>% 
  print(n = 10)


# Check model performance
folds1 <- vfold_cv(data_train1, v = 5)
data_fit_rs1 <- data_wflow1 %>% 
  fit_resamples(folds1)
collect_metrics(data_fit_rs1)

rmse(data_pred1, truth = Latitude, estimate = .pred)
rsq(data_pred1, truth = Latitude, estimate = .pred)
```


3. Exploring Trophic Level
Further exploration examined whether variables such as range size and mass might differentiate trophic groups.
A boxplot shows large variance within each trophic category, and our ROC analysis illustrates very low predictive performance (AUC close to 0.5), indicating no usable discrimination between scavengers and others based on range size.
```{r exploring trophic level, echo=FALSE}

data$Trophic.Level <- as.factor(data$Trophic.Level)

data_clean <- data %>%
    filter(!is.na(Range.Size),
           !is.na(Trophic.Level))

model_trophic_range <- lm(Range.Size ~ Trophic.Level, data = data_clean)
summary(model_trophic_range)

ggplot(data_clean, aes(x = Trophic.Level, y = Range.Size)) +
    geom_jitter()

library(tidymodels)

data$Scavenger <- factor(ifelse(data$Trophic.Level == "Scavenger", "yes", "no"),
                       levels = c("no", "yes"))

fit_scav <- logistic_reg() %>% 
  set_engine("glm") %>%
  fit(Scavenger ~ Range.Size, data = data)

data_pred <- data %>%
  mutate(.pred_yes = predict(fit_scav, data, type = "prob")$.pred_yes)

data_pred %>%
  roc_curve(Scavenger, .pred_yes, event_level = "second") %>%
  autoplot()

data_pred %>%
  roc_auc(Scavenger, .pred_yes, event_level = "second")

data_pred2 <- data_pred %>%
  mutate(Scavenger_num = if_else(Scavenger == "yes", 1, 0))

data_pred2 %>%
  yardstick::rmse(truth = Scavenger_num, estimate = .pred_yes)

data_pred2 %>%
  yardstick::rsq_trad(truth = Scavenger_num, estimate = .pred_yes)


```

#Predictve Modelling
